---
title: "rcna demo"
author: "David Oliver"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rcna demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Covarying Neighborhood Analysis Tutorial

## Minimal Working Example

### Input Data

The expected input for `rcna`'s `assocation` and `nam` functions is a list containing 
`list(samplem, obs, connectivities, samplem_key, obs_key, N)`. 

* The `samplem` element
contains a `data.frame` with sample level information (one row for each sample).
* The `obs` element contains a `data.frame` with cell level information 
e.g. `colData` of a `SingleCellExperiment` object. 
* The `connectivities` element
contains the KNN graph as a sparse matrix object. 
* The `samplem_key` and `obs_key`
elements should be character strings which identify the columns of the corresponding
element unique IDs e.g. `samplem_key = "donor_id"` and `obs_key = "cell_id"`. 
* Finally `N` should be an integer indicating the number of samples in the dataset.

`rcna` also provides `createCNAListSCE` a helper function for generating the input 
data list from a `SingleCellExperiment` as well as `associationSCE` for performing 
`rcna` directly on the `SingleCellExperiment`.

```{r}

library(rcna)
suppressPackageStartupMessages({
    library(dplyr)
    library(ggplot2)
    library(SingleCellExperiment)
    library(anndataR)
    library(ggbeeswarm)
})


demo <- system.file("ext/", "demo_data.h5ad", package = "rcna") |> 
    read_h5ad(as = "SingleCellExperiment")

demo

rcna_data <- createCNAListSCE(sce = demo, 
                              y = "case", 
                              sample.key = "id", 
                              sample.vars = c("male", "batch"), 
                              knn.graph = "connectivities")

sapply(rcna_data, class)

sapply(rcna_data$samplem, class)

# rcna expects batches and IDs to be factors
rcna_data$samplem$batch <- factor(rcna_data$samplem$batch)
rcna_data$obs$id <- factor(rcna_data$obs$id)

```

### Perform Case/Control Analysis

```{r}

res <- association(data = rcna_data, 
                   y = "case", 
                   batches = "batch", 
                   covs = "male", 
                   verbose = FALSE)

# global association p-value
res$p

# neighborhoods passed significance at FDR 5%
sum(abs(res$ncorrs) > res$fdr_5p_t)

```

#### Visualize the results

```{r}

dimplotCoords(coords = reducedDim(demo, "X_umap"), 
              y = res$ncorrs, 
              threshold = res$fdr_5p_t)

```

In the plot above, each cell is colored according to its neighborhood coefficient, 
i.e., color corresponds to the correlation of abundance in each cell's corresponding 
neighborhood to case/control status. Neighborhoods that don't pass FDR 5% are shown 
in gray. As this plot shows, cases have a greater abundance of cells from the 
left-hand cluster relative to controls.

### Analyze a More Sophisticated Phenotype

Next, perform an association test for male/female sex rather than case/control status.

```{r}

res <- association(data = rcna_data, 
                   y = "male", 
                   batches = "batch", 
                   covs = "case", 
                   verbose = FALSE)

dimplotCoords(coords = reducedDim(demo, "X_umap"), 
              y = res$ncorrs, 
              threshold = res$fdr_5p_t)

```

It's now clear that a sub-population of the left-hand cluster is more abundant in 
males. CNA automatically picked up on this without needing to be given a resolution 
parameter.

Note that it looks like some cells in the right-hand cluster pass our FDR threshold 
of 5%. However, these cells comprise only ~6% of the cells that pass the threshold 
and so are likely false discoveries.

### Relate Results to Clusters

```{r}

cldata <- data.frame(reducedDim(demo, "X_umap"), 
                     leiden = demo$leiden, 
                     ncorrs = res$ncorrs) 
colnames(cldata)[1:2] <- c("UMAP_1", "UMAP_2")

ggplot(cldata) + 
    geom_point(aes(x = UMAP_1, y = UMAP_2, color = leiden)) + 
    theme_classic()

```

We can examine within-cluster heterogeneity characterized by CNA by visualizing 
the distribution of neighborhood coefficients within each cluster.

```{r}

ggplot(cldata) +
    geom_beeswarm(aes(x = leiden, y = ncorrs, color = ncorrs), 
                              cex = 0.5) + 
    scale_color_gradient2(low = "blue", mid = "gray60", high = "red") +
    theme_classic()

```

### Relate Results to Genes

To characterize the populations driving the association, we can examine per-gene 
correlations to neighborhood coefficient. These per-gene values may be informative 
on their own, or they could be further analyzed using techniques like gene set 
enrichment analysis.

```{r}

caseCorr <- cor(as.vector(res$ncorrs), t(assay(demo, "X"))) |> 
    t() |> 
    data.frame(cor = _)

arrange(caseCorr, cor) |> tail()
arrange(caseCorr, cor) |> head()

```

<!--
## Working Directly with SingleCellExperiment Objects
-->

```{r, include=FALSE, echo=FALSE}

# demo_sce <- qs2::qs_read("~/Git/cna-rcna-test/Baella.qs2")
# 
# colPair(demo_sce, "uwot_sim") <- uwot::similarity_graph(reducedDim(demo_sce, "HARMONY"))
# 
# 
# demo <- createCNAListSCE(demo_sce, y = "treatment", sample.key = "Sample", 
#                          sample.vars = "patient_id",
#                           knn.graph = "uwot_sim")
# 
# nm <- nam(data = demo, 
#           y = "treatment", 
#           batches = "patient_id", 
#           ridge.crit = "gcv", 
#           verbose = TRUE)
# 
# res <- association(nam.result = nm, Ks = 4,
#                    return.nam = TRUE, 
#                    force.permute.all = FALSE, 
#                    allow.duplicate.perms = TRUE, 
#                    verbose = TRUE)
# res$p
# plot(res$NAM_varexp)
# plot(res$fdrs$threshold, res$fdrs$fdr, pch = 20, ylim = c(0,1))
# 
# dimplotCorr(res = res, fdr.thresh = 1)
# 
# thresh <- subset(res$fdrs, fdr < 1) |> 
#     dplyr::slice_head(n = 1) |> 
#     dplyr::pull(threshold) 
# 
# dimplotCoords(coords = reducedDim(demo_sce, "UMAP")[res$keptcells,],
#               y = res$ncorrs, 
#               label = "correlation",
#               threshold = 0)
# 
# 
# demo_sce <- associationSCE(demo_sce, 
#                            y = "Treatment", 
#                            sample.key = "orig.ident", 
#                            batches = "Sample",
#                            graph = "uwot_sim", 
#                            verbose = TRUE)
# 
# names(metadata(demo_sce)$CNA)
# 
# dimplotCoords(coords = reducedDim(demo_sce, "UMAP")[metadata(demo_sce)$CNA$keptcells,], 
#               y = metadata(demo_sce)$CNA$ncorrs, 
#               threshold = 0.3)

```




